---
title: "Predicting Barbell-Lift Execution from Wearable Sensors"
author: "Author: Sonja Sahebzad"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    code_folding: show
    fig_caption: true
    theme: cosmo
---

<style>
body { font-family: Calibri, "Segoe UI", Arial, sans-serif; color: #111; max-width: 1050px; margin: auto; line-height: 1.5; }
h1, h2, h3 { color: #111; font-weight: 600; }
table { width: auto; border-collapse: collapse; margin: 1em 0; }
th { background: #e5e5e5; } th, td { border: 1px solid #999; padding: 5px 9px; }
pre { border-left: 4px solid #777; background: #f5f5f5; }
.figure { text-align: center; } .caption { color: #333; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 7, fig.height = 4)
res <- readRDS("analysis_results.rds")
```

# Executive summary

This report predicts the five exercise-execution classes (`classe`) from accelerometers worn on the belt, forearm, arm, and dumbbell. A random forest was selected because the sensor relationships are nonlinear and may interact. Model choice was checked against a single classification tree. Five-fold cross-validation on a development set estimated random-forest accuracy at **`r sprintf('%.2f%%', 100*res$cv_rf_mean)`**; an untouched validation set produced **`r sprintf('%.2f%%', 100*res$rf_acc)`** accuracy and **`r sprintf('%.2f%%', 100*res$rf_error)`** error. The final model predicts the 20 quiz cases as shown below.

# Data and preprocessing

The training data contain `r res$n` observations. The outcome has five levels: A denotes the correct lift; B--E denote four distinct errors. I removed row identifiers, participant names, timestamps, and window bookkeeping fields. These fields describe data collection rather than body movement and could let a model memorize participants or recording order. I also removed variables observed in fewer than 95% of rows. This left `r res$p` complete sensor predictors and avoided imputing columns that are almost entirely missing.

```{r preprocessing, eval=FALSE}
training_raw <- read.csv("pml-training.csv",
  na.strings = c("NA", "#DIV/0!", ""), check.names = FALSE)
metadata <- c("X", "user_name", "raw_timestamp_part_1",
  "raw_timestamp_part_2", "cvtd_timestamp", "new_window", "num_window")
candidate <- setdiff(names(training_raw), c(metadata, "classe", ""))
predictors <- candidate[vapply(training_raw[candidate],
  function(x) mean(!is.na(x)), numeric(1)) >= 0.95]
model_data <- training_raw[c(predictors, "classe")]
model_data$classe <- factor(model_data$classe)
```

```{r class-distribution, echo=FALSE, fig.cap="Figure 1. Exercise classes are reasonably represented, supporting stratified resampling."}
cols <- gray.colors(5, start = 0.25, end = 0.80)
barplot(res$class_counts, col = cols, border = "black",
        xlab = "Exercise class", ylab = "Training observations")
```

# Validation design and candidate models

I used a fixed seed and stratified the observations by `classe`, assigning 75% to development and 25% (`r res$validation_n` rows) to final validation. All model comparison occurred within the development set. Five stratified folds provided out-of-sample predictions for every development observation. This separation prevents the final validation result from benefiting from model selection.

The candidates were (1) a classification tree, which is easy to interpret but unstable, and (2) a random forest, which averages many decorrelated trees. The forest used `mtry = floor(sqrt(52))`, a conventional classification default. The cross-validation comparison strongly favored the forest.

```{r cv-code, eval=FALSE}
for (k in 1:5) {
  rf <- randomForest(classe ~ ., data = development[fold != k, ],
    ntree = 250, mtry = floor(sqrt(length(predictors))))
  rf_prediction <- predict(rf, development[fold == k, ])
  cv_accuracy[k] <- mean(rf_prediction == development$classe[fold == k])
}
```

```{r cv-comparison, echo=FALSE, fig.cap="Figure 2. Five-fold cross-validation accuracy by candidate model."}
cv <- data.frame(
  accuracy = c(res$cv_rf, res$cv_tree),
  model = factor(rep(c("Random forest", "Single tree"), each = 5),
                 levels = c("Single tree", "Random forest")))
boxplot(accuracy ~ model, data = cv, col = c("white", "gray70"),
        border = "black", ylab = "Fold accuracy", xlab = "",
        ylim = range(cv$accuracy) + c(-0.02, 0.02))
stripchart(accuracy ~ model, data = cv, vertical = TRUE, add = TRUE,
           method = "jitter", pch = c(1, 16)[cv$model], col = "black")
```

The forest's mean cross-validation accuracy was `r sprintf('%.4f', res$cv_rf_mean)` (fold SD `r sprintf('%.4f', res$cv_rf_sd)`), versus `r sprintf('%.4f', res$cv_tree_mean)` (SD `r sprintf('%.4f', res$cv_tree_sd)`) for the single tree. Thus the substantive conclusion is sensitive to choosing a sufficiently flexible model, but the forest result itself is stable across folds.

# Final validation and expected error

After selection, I refit both candidates on the complete development set and evaluated them once on the held-out validation set. The random forest achieved `r sprintf('%.4f', res$rf_acc)` accuracy, compared with `r sprintf('%.4f', res$tree_acc)` for the single tree.

```{r confusion, echo=FALSE}
knitr::kable(res$rf_cm, caption = "Held-out confusion matrix for the selected random forest")
```

The held-out error estimate is **`r sprintf('%.2f%%', 100*res$rf_error)`**. A 95% Wilson interval places the underlying accuracy between `r sprintf('%.2f%%', 100*res$acc_ci[1])` and `r sprintf('%.2f%%', 100*res$acc_ci[2])`, equivalent to an error interval from `r sprintf('%.2f%%', 100*res$error_ci[1])` to `r sprintf('%.2f%%', 100*res$error_ci[2])`. Five-fold cross-validation implies `r sprintf('%.2f%%', 100*(1-res$cv_rf_mean))` error, while the final forest's out-of-bag estimate is `r sprintf('%.2f%%', 100*res$final_oob_error)`. These independent estimates all indicate an expected out-of-sample error below 1%, although performance on new people could be lower because the random split includes measurements from the same six participants in both sets.

```{r importance, echo=FALSE, fig.cap="Figure 3. Ten most influential predictors in the final forest."}
top <- head(res$importance_table, 10)
dotchart(rev(top$MeanDecreaseAccuracy), labels = rev(top$variable), pch = 16,
         color = "black", xlab = "Mean decrease in accuracy")
```

No manual interaction terms were added. A forest already models interactions through successive splits—for example, a forearm orientation threshold can lead to different rules depending on belt acceleration. Enumerating pairwise terms for 52 inputs would add 1,326 candidates and make a linear expansion difficult to validate. The strong, consistent resampling performance supports the automatic treatment, while the weaker single tree confirms that one simple partition is insufficient.

# Discussion

The close agreement between five-fold cross-validation, held-out validation, and out-of-bag performance supports the stability of the selected random forest. The much lower accuracy of the single tree shows that exercise quality depends on complex combinations of sensor measurements. The main limitation is that the row-level split includes observations from the same participants in development and validation. The reported error therefore estimates performance on similar participants and recording conditions; validation that holds out complete participants would provide a stricter test of generalization to new people.

# Predictions for the 20 test cases

The selected random forest was refit to all `r res$n` labeled rows with 750 trees. The following are the requested predictions, ordered by `problem_id`.

```{r predictions, echo=FALSE}
knitr::kable(res$predictions, col.names = c("Problem ID", "Predicted classe"))
```

# Conclusion

The random forest was the strongest candidate, achieving **`r sprintf('%.2f%%', 100*res$rf_acc)`** accuracy on the untouched validation set and an estimated out-of-sample error of **`r sprintf('%.2f%%', 100*res$rf_error)`**. Its consistent cross-validation and out-of-bag results support the final choice. After refitting on all labeled observations, the model generated predictions for each of the 20 required test cases.

# Reproducibility and source

This HTML document was knitted from the accompanying R Markdown file. The complete fitting script, fixed random seeds, source CSV files, saved results, and quiz-output files are included in the project directory. Package and session versions appear below.

```{r session-info}
sessionInfo()
```

The data were provided by the Pontifical Catholic University of Rio de Janeiro's [Weight Lifting Exercise Dataset](http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har). The course copies of the [training data](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv) and [20 test cases](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv) are retained unchanged.


